---
title: "Count data preprocessing"  
author: "Jule Freudenthal"  
date: "`r Sys.Date()`"  
output: github_document
always_allow_html: yes
---
```{r setup, include=FALSE}
# Install packages
if (!require("knitr")) install.packages("knitr")
if (!require("rprojroot")) install.packages("rprojroot")

# Set global options
knitr::opts_chunk$set(echo = TRUE)

# Set global working directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

**R version:** 4.3.1 (2023-06-16), Beagle Scouts  
**Packages**  

* dplyr v. 1.1.4  
* tidyr v. 1.3.1  

```{r, message=FALSE, warning=FALSE}
# Install packages
if (!require("dplyr")) install.packages("dplyr")
if (!require("tidyr")) install.packages("tidyr")

# Load packages
library(dplyr)
library(tidyr)
```

## 01Combine PR<sup>2</sup> and SILVA tables

First, we adjusted some taxonomic information. 

### 01.1 PR<sup>2</sup> 
```{r, echo=FALSE, results='hide'}
# Create new directory for the count data
if(!file.exists("Data/PreprocessedData")){dir.create("Data/PreprocessedData")}
```

```{r, message=FALSE}
# Load PR2 data
PR2=read.csv("Data/OriginalData/PR2CountTable.txt", header = TRUE, sep = "\t", dec = ".")

# Delete Filosa from Cercozoa
PR2$Class <- sub("^Filosa-","", PR2$Class)
PR2$Order <- sub("^Filosa-","", PR2$Order)
PR2$Family <- sub("^Filosa-","", PR2$Family)
PR2$Genus <- sub("^Filosa-","", PR2$Genus)

# Delete numbers from genera
PR2[grepl("[a-z][1-9]$|[a-z][1-9][1-9]$", PR2$Genus) &
      !grepl("clade", PR2$Genus) &
      !grepl("Sarcocystidae1", PR2$Genus),"Genus"] <- 
  sub("[1-9]$|[1-9][1-9]$","",
      PR2[grepl("[a-z][1-9]$|[a-z][1-9][1-9]$", PR2$Genus) &
            !grepl("clade", PR2$Genus) &
            !grepl("Sarcocystidae1", PR2$Genus),"Genus"])
PR2[grepl("_[1-9]$", PR2$Genus) &
      !grepl("Labyrinthulaceae_ANT10_3", PR2$Genus),"Genus"] <-
  sub("_[1-9]$", "", PR2[grepl("_[1-9]$|_[1-9][1-9]$", PR2$Genus) &
                           !grepl("Labyrinthulaceae_ANT10_3", PR2$Genus),"Genus"])
PR2[grepl("Babesia",PR2$Genus),"Genus"] <- "Babesia"
PR2[grepl("Sarcocystis",PR2$Genus),"Genus"] <- "Sarcocystis"
PR2[grepl("Cryptosporidium",PR2$Genus),"Genus"] <- "Cryptosporidium"
PR2[grepl("Piroplasmorida1",PR2$Genus),"Genus"] <- "Piroplasmorida1_X"
PR2[grepl("Piroplasmorida6",PR2$Genus),"Genus"] <- "Piroplasmorida6_X"

# Fix some taxonomic ranks
PR2[PR2$Genus %in% c("Arcuospathidium", "Epispathidium", "Spathidium"),"Order"] <- "Litostomatea_X"
PR2[PR2$Genus %in% c("Arcuospathidium", "Epispathidium", "Spathidium"),"Family"] <- "Litostomatea_XX"
PR2[PR2$Genus == "Balantidium","Family"] <- "Trichostomatia_X"
PR2[PR2$Genus == "Cyclidium","Order"] <- "Oligohymenophorea_X"
PR2[PR2$Genus == "Cyclidium","Family"] <- "Oligohymenophorea_XX"
PR2[PR2$Genus == "Frontonia","Family"] <- "Peniculia_X"
PR2[PR2$Genus == "Zoothamnium","Family"] <- "Peritrichia_2_X"
PR2[PR2$Genus == "Physarum","Family"] <- "Physaraceae"
PR2[PR2$Genus == "Pleuronema","Family"] <- "Scuticociliatia_2_X"

# Summarise data by genus
PR2 <- data.frame(PR2 %>% mutate_if(is.factor, as.character))
PR2 <- data.frame(PR2 %>% 
                    group_by(across(where(is.character))) %>% 
                    summarise(across(everything(), sum)))
```

### 01.2 SILVA  

```{r, message=FALSE}
# Load SILVA data
SILVA=read.csv("Data/OriginalData/SILVACountTable.txt", header = TRUE, sep = "\t", dec = ".")

# Replace NAs
SILVA[is.na(SILVA)] <- ""

# Replace uncultured taxa
for(i in 2:6){
  
  # Get names
  OrigName <- colnames(SILVA)[i]
  
  # Get indices
  indices <- 
    grep("uncultured$|uncultured_actinobacterium$|uncultured_actinomycete$|
          |uncultured_alga$|uncultured_anaerobic_ammonium-oxidizing_bacterium$|
          |uncultured_anaerobic_bacterium$|uncultured_archaeon$|
          |uncultured_bacterium$|uncultured_beta_proteobacterium$|
          |uncultured_compost_bacterium$|uncultured_crenarchaeote$|
          |uncultured_cyanobacterium$|uncultured_Cyanobacterium_sp$|
          |uncultured_deep-sea_bacterium$|uncultured_delta_proteobacterium$|
          |uncultured_diatom$|uncultured_endolithic_bacterium$|
          |uncultured_eukaryote$|uncultured_forest_soil_bacterium$|
          |uncultured_gamma_proteobacterium$|uncultured_haloalkaliphilic_bacterium$|
          |uncultured_iron-reducing_bacterium$|uncultured_marine_bacterium$|
          |uncultured_marine_cyanobacterium$|uncultured_marine_eukaryote$|
          |uncultured_marine_microorganism$|uncultured_marine_rappemonad$|
          |uncultured_methanogenic_archaeon$|uncultured_microorganism$|
          |uncultured_organism$|uncultured_phototrophic_eukaryote$|
          |uncultured_planctomycete$|uncultured_prokaryote$|
          |uncultured_proteobacterium$|uncultured_rumen_bacterium$|
          |uncultured_sludge_bacterium$|uncultured_soil_archaeon$|
          |uncultured_soil_bacterium$|uncultured_sediment_bacterium$|
          |uncultured_sulfur-oxidizing_symbiont_bacterium$|
          |uncultured_thaumarchaeote$|uncultured_thermal_soil_bacterium$", SILVA[,OrigName])
  
  # Get names of high taxonomic level
  NewName <- colnames(SILVA)[i-1]
  SILVA[indices,OrigName] <- ifelse(grepl("_X+$", SILVA[indices,NewName]),
                                    paste0(SILVA[indices,NewName], "X"),
                                    paste0(SILVA[indices,NewName], "_X"))
}

# Replace empty columns
for(i in 2:6){
  
  # Get names
  OrigName <- colnames(SILVA)[i]
  
  # Get indices
  indices <- which(SILVA[,OrigName]=="")
  
  # Get names of high taxonomic level
  NewName <- colnames(SILVA)[i-1]
  SILVA[indices,OrigName] <- ifelse(grepl("_X+$", SILVA[indices,NewName]),
                                    paste0(SILVA[indices,NewName], "X"),
                                    paste0(SILVA[indices,NewName], "_X"))
}

# Summarise by genus
SILVA <- data.frame(SILVA %>% mutate_if(is.factor, as.character))
SILVA <- data.frame(SILVA %>% 
                      group_by(across(where(is.character))) %>% 
                      summarise(across(everything(), sum)))
```

### 01.3 Filter data to exclude cholorplast sequences, land plants and macroscopic animals

We filtered for putative contaminants, such as sequences derived from macroscopic animals and land plants (Embryophyceae). Further, we exclude chloroplast sequences from the data set.

```{r}
# Exclude cholorplast sequences from SILVA data
SILVA <- SILVA[SILVA$Order != "Chloroplast",]

# Filter to exclude land plants non microscopic metazoa
PR2 <- PR2[PR2$Class != "Embryophyceae",]

# Filter to exclude non microscopic metazoa
PR2 <- 
  PR2[!(PR2$Phylum == "Metazoa" & 
          (!PR2$Class %in% c("Gastrotricha", "Nematoda", "Rotifera", "Tardigrada"))),]
```

### 01.4 Combine tables  

We combine the PR<sup>2</sup> and SILVA data and assign a new column differentiating between Prokaryota, Algae, heterotrophic Protists, Fungi and Microscopic Metazoa. Further, we create one table for count-, meta- and taxonomy data respectively. We transform the data so that for  

1. **Count data** one row contains all information to one taxon for each sample. Column names must be sample-IDs and row names taxa-IDs.  
2. **Metadata** One column per metadata factor providing information for each sample. Column names are factors, row names are sample-IDs.  
3. **Taxonomy** Each column is one taxonomic rank, providing the classification of each taxon. Column names are taxonomic ranks, row names are taxa-IDs. 

```{r}
# Add a new column 'Super_Domain' to the Silva data
SILVA$Super_Domain <- "Prokaryota"

#------------ Add a column 'MicrobialCommunity'
# Prokaryota
SILVA$MicrobialCommunity <- "Prokaryota"

# Heterotrophic Protists
PR2$MicrobialCommunity <- "Heterotrophic Protists"

# Algae
PR2[PR2$Domain == "Archaeplastida" & PR2$Phylum %in% 
      c("Chlorophyta", "Glaucophyta", "Prasinodermophyta", "Rhodophyta"),
    "MicrobialCommunity"] <- "Algae"
PR2[PR2$Phylum == "Streptophyta" & PR2$Class %in% 
      c("Charophyceae", "Chlorokybophyceae", "Coleochaetophyceae", "Klebsormidiophyceae", 
        "Mesostigmatophyceae", "Zygnemophyceae"), "MicrobialCommunity"] <- "Algae"
PR2[PR2$Domain == "Hacrobia" & PR2$Phylum %in% 
      c("Cryptophyta", "Haptophyta"), "MicrobialCommunity"] <- "Algae"
PR2[PR2$Phylum == "Ochrophyta", "MicrobialCommunity"] <- "Algae"
PR2[PR2$Order == "Euglenophyceae", "MicrobialCommunity"] <- "Algae"
PR2[PR2$Genus %in% c("Vitrella", "Chromera", "Paulinella"), "MicrobialCommunity"] <- "Algae"

# Microscopic Metazoa
PR2[PR2$Phylum=="Metazoa","MicrobialCommunity"] <- "Microscopic Metazoa"

# Fungi
PR2[PR2$Phylum=="Fungi","MicrobialCommunity"] <- "Fungi"

# Eukaryota_X
PR2[PR2$Domain=="Eukaryota_X","MicrobialCommunity"] <- "Eukaryota_X"

# Combine both tables
CountData <- rbind(PR2, SILVA)

# Shorten column names
colnames(CountData) <- sub("Tilia_CanopyBark_0", "TB", 
                           sub("Quercus_CanopyBark_0", "QB", 
                               sub("Acer_CanopyBark_0", "AB", colnames(CountData))))

# Assign row names
rownames(CountData) <- paste0("Genus",sprintf('%0.4d', seq(nrow(CountData))))

# Separate data in count and taxonomy data
CountData <- data.frame(CountData %>% mutate_if(is.factor, as.character))
counts <- data.frame(select_if(CountData, is.numeric))
taxonomy <- data.frame(select_if(CountData, is.character))

# Load metadata, shorten Sample IDs and assign Sample ID as row names
metadata <- read.csv("Data/OriginalData/Metadata.txt", header = TRUE, sep = "\t", dec = ".")
metadata$SampleID <- sub("Tilia_CanopyBark_0", "TB", 
                         sub("Quercus_CanopyBark_0", "QB", 
                             sub("Acer_CanopyBark_0", "AB", metadata$SampleID)))
rownames(metadata) <- metadata$SampleID

# Check if corresponding row and column names match, else save tables as Data
if(!all(identical(colnames(counts), rownames(metadata)), identical(rownames(counts), rownames(taxonomy)))){
  stop("Corresponding row and column names must match")
} else {
  # Save files
  save(counts, metadata, taxonomy, file = "Data/PreprocessedData/01CanopyBarkData.rds")
}
```
```{r, echo=FALSE, results='hide'}
# Clear console and workspace
cat("\014") 
rm(list=ls())
```

## 02 Filter for low abundant data

We exclude genera with proportional reads below 0.001 % across all samples. 

```{r, results='hold'}
# Load RNA count- and meta- and taxonomy data
load("Data/PreprocessedData/01CanopyBarkData.rds")

# Filter for most abundant
filter <- 0.00001
print(paste0(sum(rowSums(counts) >= sum(counts)*filter), " out of ", nrow(counts), 
             " genera are kept for further analysis. Only ",
             (sum(counts)-sum(counts[rowSums(counts) >= 
                                       sum(counts)*filter,]))/sum(counts)*100 ,
             "% of the reads are excluded."))
counts <- counts[rowSums(counts) >= sum(counts)*filter,]

# Adjust taxonomy
taxonomy <- taxonomy[rownames(taxonomy) %in% rownames(counts),]

# Check if corresponding row and column names match, else save tables as Data
if(!all(identical(colnames(counts), rownames(metadata)), 
        identical(rownames(counts), rownames(taxonomy)))){
  stop("Corresponding row and column names must match")
} else {
  rownames(taxonomy) <- paste0("Genus",sprintf('%0.4d', seq(nrow(taxonomy))))
  rownames(counts) <- rownames(taxonomy)
  save(counts, metadata, taxonomy, 
       file = "Data/PreprocessedData/02CanopyBarkDataFiltered.rds")
}

```
```{r, echo=FALSE, results='hide'}
# Clear console and workspace
cat("\014") 
rm(list=ls())
```

## 02 Assign Traits

We assign functional traits to the taxa identified, using published reference databases. Fungi were classified into functional groups (lichen-forming fungi, (facultative) yeasts, and plant parasites) according to [Nguyen et al., 2016](http://www.scopus.com/inward/record.url?scp=84959902395&partnerID=8YFLogxK).
Algae were classifiedaccording to [Sanders & Masumoto, 2021](https://www.cambridge.org/core/journals/lichenologist/article/lichen-algae-the-photosynthetic-partners-in-lichen-symbioses/A4808B98986967A742EF0DCFF187A937) as lichen-forming algae. 

```{r}
# Load RNA count- and meta- and taxonomy data
load("Data/PreprocessedData/02CanopyBarkDataFiltered.rds")

# Assign Genus ID as column
taxonomy$ID <- rownames(taxonomy)

#------------- Fungi
# Load fungi traits
fungi <- read.table("Data/OriginalData/Traits/Fungi.txt", sep = "\t", header = T)

# Separate taxonomy
fungi <- fungi %>% separate(taxonomy,c("Class", "Order", "Family", "Genus"), sep = ";")

# Merge with taxonomy
fungi <- 
  merge(taxonomy[taxonomy$MicrobialCommunity=="Fungi",],
        fungi[,c("Class", "Order", "Family", "Genus", "Feeding", "Lifestyle", "Growth.Morphology")], 
        all.x = T, all.y = F)

# Keep only traits for lichen, pathogenes and growth morphology
fungi[!fungi$Lifestyle %in% c("Lichenized", "Plant Pathogen"),"Lifestyle"] <- NA
fungi[is.na(fungi$Growth.Morphology) | fungi$Growth.Morphology == "NULL","Growth.Morphology"] <- NA

# Assign a column for lichen
fungi$Lichen[fungi$Lifestyle == "Lichenized"] <- "Mycobiont"

# Merge with taxonomy
taxonomy <- merge(taxonomy[taxonomy$MicrobialCommunity!="Fungi",], fungi, all=T)

#------------- Photobionts
# Load photobionts traits
photobionts <- read.table("Data/OriginalData/Traits/Photobionts.txt", sep = "\t", header = T)

# Merge with taxonomy
photobionts <- 
  merge(taxonomy[taxonomy$MicrobialCommunity == "Algae" | taxonomy$Phylum == "Cyanobacteria",
                 c("Super_Domain", "Domain", "Genus", "ID")], photobionts, all = F)

# Merge with taxonomy table
taxonomy$Lichen[taxonomy$ID %in% photobionts$ID] <- "Photobiont"

# Assign ID as rownames
rownames(taxonomy) <- taxonomy$ID
taxonomy$ID <- NULL

# Check if corresponding row and column names match, else save tables as Data
taxonomy <- taxonomy[match(rownames(counts),rownames(taxonomy)),]
if(!all(identical(colnames(counts), rownames(metadata)), 
        identical(rownames(counts), rownames(taxonomy)))){
  stop("Corresponding row and column names must match")
} else {
  # Save files
  save(counts, metadata, taxonomy, file = "Data/PreprocessedData/03CanopyBarkDataTraitsFiltered.rds")
}
```
```{r, echo=FALSE, results='hide'}
# Clear console and workspace
cat("\014") 
rm(list=ls())
```

